Free surface computations in mantle convection models

### 2016-04-04 free-surface-paper.tex
Minor edits.

\section{Governing Equations}
\label{sec:governing}

@@

Armed with the normal modes and relaxation times of the Stokes system, we can write down the
formal solution to Equation~\eqref{eq:surface_evolution}. Let the initial surface topography be 
represented by a linear combination of its normal modes
\begin{equation}
\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}),
\end{equation}
then the time evolution of the free surface is given by
\begin{equation}
\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
\end{equation}
Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
the solution and surface topography into its normal modes and integrate those separately. 
Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
The normal mode solution is instructive, however. 
Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
The decay equation is the archetypical example of a stiff ordinary differential equation.
If we were to numerically integrate this in time with a forward Euler method, we would find the 
time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
\begin{equation}
\Delta t  \le 2 \tau_{\mathrm{min}},
\label{eq:cfl_euler}
\end{equation}
where $\tau_{\mathrm{min}}$ is this minumum decay time.
The maximum allowable timestep is limited by the minimum relaxation timescale.
If a larger timestep than this is taken then those modes will go unstable.
The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
and so it will be those which go unstable first, a phenomenon which has been called 
a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.

@@

\section{Numerical Results}
\label{sec:results}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/amr.pdf}
\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement. The timestep $\Delta t$ is 500 yr. We compare the maximum topography at 3 Myr with its value at high resolution ($\sim$398 m). For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients. Quasi-implicit and nonstandard finite difference timestepping converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}
\label{fig:amr}
\end{figure*}

>>>>

\section{Governing equations}
\label{sec:governing}

@@

Armed with the normal modes and relaxation times of the Stokes system, we can write down the
formal solution to Equation~\eqref{eq:surface_evolution}. Let the initial surface topography be 
represented by a linear combination of its normal modes
\begin{equation}
\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}),
\end{equation}
then the time evolution of the free surface is given by
\begin{equation}
\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
\end{equation}
Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
the solution and surface topography into its normal modes and integrate those separately. 
Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
The normal mode solution is instructive, however. 
Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
The decay equation is the archetypical example of a stiff ordinary differential equation.
If we were to numerically integrate this in time with a forward Euler method, we would find the 
time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
\begin{equation}
\Delta t  \le 2 \tau_{\mathrm{min}},
\label{eq:cfl_euler}
\end{equation}
where $\tau_{\mathrm{min}}$ is this minumum decay time.
The maximum allowable timestep is limited by the minimum relaxation timescale.
If a larger timestep than this is taken then those modes will become unstable.
The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
and so it will be those which become unstable first, a phenomenon which has been called 
a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.

@@

\section{Numerical results}
\label{sec:results}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/amr.pdf}
\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement. The timestep $\Delta t$ is 500 yr. We compare the maximum topography at 3 Myr with its value at high resolution ($\sim$398 m). For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients. Quasi-implicit (QI) and nonstandard finite difference (NSFD) timestepping converge similarly, but the adaptive mesh refinement simulations require significantly fewer unknowns.}
\label{fig:amr}
\end{figure*}
